Exploratory Analysis of Prosper Loan Data by Sidhant Karamchandani
===================================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(memisc)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}
# Load the Data
loan_data = read.csv("prosperLoanData.csv")
```

# Introduction

In this project, I'll be exploring loan data provided by the lender Prosper.
They have publicly provided laon data for a variety of loan types, from the year
2005 to 2014. As harmful as the 2008 recession was, the reasons it formed are
fascinating and we can learn from them. The purpose of this project is to learn
a bit more about how Americans behave when it comes to personal finance and 
taking loans.


# Univariate Plots Section

Let's start with learning the basics of the dataset itself.

```{r echo=FALSE, Univariate_Plots}
dim(loan_data)
str(loan_data)
```

We have 113937 loan observations, with 81 various variables relating to the loan
or borrower characteristics. The LoanOrigination dates are stored as factored 
strings, though. I want to convert these to date objects.

```{r}
# Converting date column from factored strings to date objects
loan_data$LoanOriginationDate <- as.Date(loan_data$LoanOriginationDate)
str(loan_data$LoanOriginationDate)
range(loan_data$LoanOriginationDate)
```

Perfect. Now I know my data is from 2005 to 2014. What kind of loans were taken
out during this time period? First let's add a column to view loan types by
their name, not number.

```{r}
# New column to display loan type names
loan_data$ListingCategory <- factor(loan_data$ListingCategory..numeric., labels = c("Not Available", "Debt Consolidation", "Home Improvement", "Business", "Personal Loan", "Student Use", "Auto", "Other", "Baby&Adoption", "Boat", "Cosmetic Procedure", "Engagement Ring", "Green Loans", "Household Expenses", "Large Purchases", "Medical/Dental", "Motorcycle", "RV", "Taxes", "Vacation", "Wedding Loans"))
```
```{r echo=FALSE, message=FALSE, warning=FALSE}
# Histogram of loan types
ggplot(aes(x = ListingCategory), data = loan_data) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
```

It seems that a LOT of people are taking out loans to pay off other debt... Is
this worrisome? We'll explore that later.


Even though Student Use above is a relatively low frequency loan, some people 
are worried that high student loan debt could create a future recession. Let's 
leave out all other occupations and see what kinds of students are taking loans.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Histogram of student loans after subsetting for only those Occupations
students <- loan_data[grepl("*Student*", loan_data$Occupation), ]

ggplot(aes(x = Occupation), data = students) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
```

The highest frequency loan taken by a student is a loan taken by a graduate
student. This makes sense, as graduate school costs far more than undergraduate
school on average, so a grad student might be more in need of immediate cash.

Next let's look at loan sizes.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Histogram of loan amounts
ggplot(aes(x = LoanOriginalAmount), data = loan_data) +
  geom_histogram(binwidth = 1000, color = "white") +
  scale_x_continuous(breaks = seq(0, 36000, 1000), limits = c(0, 36000)) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
```

The loan amounts are not exactly normally distributed. They are skewed to the
right, with a couple outliers beyond $30,000. The median appears to be around
the high 4000s mark. Let's dive deeper into the measures of center:

```{r echo=FALSE, message=FALSE, warning=FALSE}
# loan amount centers
summary(loan_data$LoanOriginalAmount)
```
I was wrong. The median is far higher than I thought. The mean is also much 
higher than my estimated median, but that makes sense because of the relatively
huge outliers and spikes around the 10,000 and 15,000 mark. I also want to know
more about the lengths of the loans, and their interest rates.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Histogram of loan terms
ggplot(aes(x = Term), data = loan_data) +
  geom_histogram(color = "white")
```

Not much variability here...let's check a table:

```{r echo=FALSE, message=FALSE, warning=FALSE}
# table of loan terms
table(loan_data$Term)
```

There's only three values for loan term: 1 year, 3 years, and 5 years. The 
majority are 3 years, followed by 5 years, and then 1 year. Next up, interest 
rates:

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Histogram of interest rate
ggplot(aes(x = loan_data$LenderYield), data = loan_data) +
  geom_histogram(binwidth = 0.01, color = "white") +
  scale_x_continuous(breaks = seq(0, 0.4, 0.02), limits = c(0, 0.4)) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
```

Interest rates are much more varied. First of all, interest rates are normally
distributed, with a center around 0.16 - 0.17. There's another spike around
0.31. Let's see if my assumptions are right:

```{r echo=FALSE, message=FALSE, warning=FALSE}
# summary of interest rates
summary(loan_data$LenderYield)
```

I do not understand how an interest rate can be negative. That data point must
be incorrect. But, I was pretty spot on with my assumptions about the center.

Now that I know the loan amounts, their lengths, and their interest rates, what 
are the incomes of the borrowers?

```{r echo=FALSE, message=FALSE, warning=FALSE}
# ordering the income ranges first
loan_data$IncomeRange <- ordered(loan_data$IncomeRange, levels = 
                                   c("Not displayed", "Not employed", "$0",
                                     "$1-24,999", "$25,000-49,999",
                                     "$50,000-74,999", "$75,000-99,999",
                                     "$100,000+"))
# income ranges
ggplot(aes(x = IncomeRange), data = loan_data) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
```

Borrower's incomes seem to be somewhat normally distributed. By looking at this 
plot I'd guess the interquartile range is 25K - 75K. I want to take this 
further. What is each borrower's cash flow? Meaning their monthly income minus
their monthly loan payment.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# new cash flow variable
loan_data$CashFlow <- 
  (loan_data$StatedMonthlyIncome - loan_data$MonthlyLoanPayment)

ggplot(aes(x = CashFlow), data = loan_data) +
  geom_histogram(binwidth = 200, color = "white") +
  scale_x_continuous(breaks = seq(0, 10000, 200), limits = c(0, 10000)) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
```

Cash Flow is normally distributed. Let's explore a summary to find outliers and
the measures of center. 

```{r echo=FALSE}
# cash flow
summary(loan_data$CashFlow)
```

The measures of center lie around 4000-5000, which is a healthy monthly cash 
flow. Some people have negative cash flows, which could have harmful financial 
consequences if not addressed. There are some borrowers with massive cash flows,
upto 1.7 million. These must be business owners and executives right?

```{r echo=FALSE}
# cash flow > 1 million
with(subset(loan_data, CashFlow > 1000000), table(Occupation))
```

The person is listed as a "Professional". Definitely an outlier. Okay, how much 
are borrowers in debt?

```{r echo=FALSE, message=FALSE, warning=FALSE}
# borrower debt to income ratios
ggplot(aes(x = DebtToIncomeRatio), data = loan_data) +
  geom_histogram(binwidth = 0.1, color = "white")
```

Debt to income ratio is normally distributed, with a couple outliers around 
10.0. For the most part, borrowers do not have massive debt compared to their
income, except for the high outliers. I'll also add another quick plot without
the high outlier:

```{r echo=FALSE, message=FALSE, warning=FALSE}
# borrower debt to income ratios
ggplot(aes(x = DebtToIncomeRatio), data = loan_data) +
  geom_histogram(binwidth = 0.1, color = "white") +
  scale_x_continuous(limits = c(0, 1.5))
```

Now we have a much better look at DebtToIncomeRatio. It's still mostly 
normally distributed, with a bit of a skew to the right.

Someone once said, "a friend in need is a friend indeed". How much financial
help are borrowers recieving from friends?

```{r echo=FALSE, message=FALSE, warning=FALSE}
# investment from friends amounts
ggplot(aes(x = InvestmentFromFriendsAmount), data = loan_data) +
  geom_histogram(color = "white")
```

Seems odd to just see one spike in the beginning of the x-axis, but let's take
a closer look:

```{r echo=FALSE}
# investment from friends amounts
summary(loan_data$InvestmentFromFriendsAmount)
```

It seems that most people do not take money from friends; and if they do, it's
in very small amounts. The mean is only $16.55. But there also is a a very high
outlier at $25,000. Perhaps this person has wealthy friends, or maybe the
friends were acting as angel investors for a business. I will create the same
plot again below, but without values of 0.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# investment from friends amounts
filtered_data <- subset(loan_data, InvestmentFromFriendsAmount != 0)
ggplot(aes(x = InvestmentFromFriendsAmount), data = filtered_data) +
  geom_histogram(color = "white") +
  scale_x_continuous(breaks = seq(100, 25100, 1000), limits = c(100, 25100)) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
```

This visualization looks a lot better. Most investments from friends are
relatively low; this data is very skewed to the right.

Next, I'll explore other investors that are not friends. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
# investors
ggplot(aes(x = Investors), data = loan_data) +
  geom_histogram(color = "white") +
  scale_x_continuous(breaks = seq(100, 1300, 200), limits = c(100, 1200)) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
```

The number of investors per loan is also very skewed to the right, with a huge
spike around 50-70. I'm guessing this is probably because most requested loans
are at or below $10,000, as we saw above. Only a minority of loans are above
that amount, meaning most loans probably don't need a huge number of investors
to cover it.

Lastly, let's look at ProsperRatings (essentially calculated ratings assessing
the quality of the loan & borrower ability) assigned by Prosper, and loan 
statuses to see how loans actually turned out.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# prosper ratings
loan_data$ProsperRating..Alpha. <- ordered(loan_data$ProsperRating..Alpha., 
                                           levels = c("", "AA", "A", "B", "C", 
                                                      "D", "E", "HR"))

ggplot(aes(x = loan_data$ProsperRating..Alpha.), data = loan_data) +
  geom_bar(color = "white", fill = "#4286f4")
```

There are a bunch of loans without a rating, which is okay as this was specified
in the dataset description. But otherwise, prosper ratings are actually very
normally distributed!

```{r echo=FALSE}
# loan statuses
ggplot(aes(x = LoanStatus), data = loan_data) +
  geom_bar(color = "white") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
```

Only a relatively small proportion of loans are defaulted or past due. This is
very reassuring!


# Univariate Analysis

### What is the structure of your dataset?
There are 113,937 loan observations, with 82 variables being tracked. The data
is from 2005 to 2014. The data tracks various aspects of the loan itself, and of
the borrower.

### What is/are the main feature(s) of interest in your dataset?
The main factors I'll be exploring going forward will involve defaulted loans. I
want to see if it is possible to predict whether a loan will be defaulted, based
on borrower & loan characteristics.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?
Other features that will support my analysis include: loan types, loan amounts, 
credit/risk scores, late payments, defaulted loans, incomes, and debt. 

### Did you create any new variables from existing variables in the dataset?
Yes, I created the CashFlow variable, which is StatedMonthlyIncome minus
MonthlyLoanPayment.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?
Loan amounts are skewed to the right. IncomeRange is somewhat normally
distributed. Debt to Income ratios are also normally distributed, with a couple
very high outliers. Investments from friends are extremely low except for one
anomaly. Cash Flow is normally distributed with a couple outliers.
I changed LoanOriginationDate from a factor to a date format, and I changed 
IncomeRange to an ordered factor. A date variable makes no sense to analyze
unless it is actually presented as a date. And the plot for IncomeRange was also
hard to make sense of unless it was ordered. 


# Bivariate Plots Section

After the housing market crash of 2008, I would assume that financial
institutions are much better now at assessing qualifications for loans. Let's 
see how accurate Prosper is. I'll look at loan statuses and their associated
ratings at the time of the loan creation.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots}
# Loan statuses vs Prosper ratings
ggplot(aes(x = LoanStatus), data = loan_data) +
  geom_histogram(aes(fill = loan_data$ProsperRating..Alpha.), 
                 position = "dodge", stat = "count") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
```

First of all, a lot of loans have a rating of N/A because that data is only
available after July 2009. For loans that were completely paid in full, Prosper
was not very accurate with their scores. For completed loans, Prosper had given
a similar amount of good ratings (AA, A, B), as bad ratings (C, D, E). Defaulted
loans are hard to see, so let's zoom in to those below.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Loan statuses vs Prosper ratings
defaulted = subset(loan_data, 
                   LoanStatus == 'Defaulted' & ProsperRating..Alpha. != '')
ggplot(aes(x = LoanStatus), data = defaulted) +
  geom_bar(aes(fill = defaulted$ProsperRating..Alpha.), 
                 position = "dodge", stat = "count")
```

For defaulted loans, Prosper was fortunately more accurate in their pre-ratings.
The highest rating, AA, was given the least. Bad ratings (C, D, E) were given in
much higher proportions than the best 3 ratings.

So it seems Prosper takes the conservative route and hands out bad ratings at
the slighest notion of a potential defaulted loan in the future. How about the
individual investors who fund these loans? Let's see how investors act based on
debt to income ratios.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Debt to Income ratios vs number of investors lending
ggplot(aes(x = DebtToIncomeRatio, y = Investors), 
       data = loan_data) +
  geom_point(color = "#109876", alpha = 0.33, size = 1.0, position = "jitter") +
  scale_x_continuous(breaks = seq(0, 1.0, 0.05), limits = c(0, 1.0)) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
```

Investors also seem to pretty careful. The higher a borrower's debt to income
ratio, the lower the number of investors willing to lend. Let's see if there is
any correlation between the two.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Debt to Income ratios vs number of investors lending. CORRELATION COEFFICIENT
cor.test(x = loan_data$DebtToIncomeRatio, y = loan_data$Investors)
```

R is not able to find a significant correlation coefficient. Perhaps the
presence of outliers, or lack of normality threw the results off. But based on 
the plot above, I still stick with my conclusion.

All is well and good for the investor/lender when the borrower provides their
income, but what about when the income isn't verifiable? Is this usually
associated with an inability to pay off a loan? Or is it just random 
circumstance? Let's see.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Loan Statuses vs Verifiable Income
ggplot(aes(x = LoanStatus), data = loan_data) +
  geom_histogram(aes(fill = loan_data$IncomeVerifiable), 
                 position = "dodge", stat = "count") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
```

There doesn't seem to be a significant association between an income being 
unverifiable and the borrower not being able to pay off their loan.

The average citizen has a variety of bill payments due each month. How can 
investors/lenders have peace of mind that their loan will be paid, based on a
borrower's monthly cash flow?

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Loan Status vs Cash Flow
ggplot(aes(x = LoanStatus, y = CashFlow), data = loan_data) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  coord_cartesian(ylim = c(0, 10000))
```

Not too surprisingly, the range and median cash flows for completed loans are
higher than the range and median cash flows of defaulted loans. That being said,
the differences are not drastic. Also, the spread and center of cash flows for
late payments are no different than those of completed loans, meaning cash flow
is not a good predictor of whether a borrower will pay late. 

Let's see if investors are more confident with higher calculated cash flows.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Investors vs Cash Flow
# using "summary_bin" to break my cash flow into bins of 500, otherwise the plot
# is far too overplotted and looks like a mess. Then in each bin, I'm using
# fun.y to calculate the median number of Investors per bin. summary_bin is 
# similar to summary, except the fun.y is applied to the y values of BINS of
# x values, instead of every single x value
ggplot(aes(x = CashFlow, y = Investors), data = loan_data) +
  geom_point(stat = 'summary_bin', fun.y = median, binwidth = 500) +
  geom_smooth(stat = 'summary_bin', fun.y = median, binwidth = 500, 
              size = 0.5) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  ylab("Median Investors") +
  scale_x_continuous(breaks = seq(0, 10000, 500), limits = c(0, 10000))
```

It looks like there is a positive correlation here. Let's confirm with
a correlation test:

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Investors vs Cash Flow
cor.test(x = loan_data$CashFlow, y = loan_data$Investors)
```

The calculated correlation coefficient is pretty weak, but that is probably
because I did not calculate it using Cash Flow in bins, nor Investors as a 
median of those bins. 

Continuing on, how do investors act based upon ProsperScore (risk score) ?

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Investors vs Prosper Score
ggplot(aes(x = ProsperScore, y = Investors), data = loan_data) +
  geom_point(stat = 'summary', fun.y = median) +
  geom_smooth(stat = 'summary', fun.y = median, size = 0.5) +
  ylab("Median Investors") +
  scale_x_continuous(breaks = seq(0, 10, 1), limits = c(0, 10))
```

Once again, there seems to be a mostly positive correlation between the median
number of Investors according to ProsperScore. As ProsperScore increases
(meaning a better score), the median number of Investors increases. Confirming
with correlation test:

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Investors vs Cash Flow
cor.test(x = loan_data$ProsperScore, y = loan_data$Investors)
```

There's definitely more of a positive correlation here. The only reason I think
the correlation coefficient isn't higher is because of the unexplained drop in
investors at ProsperScores between 2-4. There must be other factors that
investors analyzed to explain that drop, or maybe there weren't many loans in
that range of scores so the median number of investors is low.


# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?
For loans that ended up fully completed, Prosper wasn't super accurate with
their ProsperRatings. There were almost as many bad ratings as good ratings. But
conversely, for loans that ended up Defaulted, Prosper was a lot more accurate,
dishing out far more bad ratings. This means Prosper is pretty conservative in
their ratings. The spread and center of borrower Cash Flow for completed loans
were visibly higher than those of defaulted loans, but not significantly 
different from loans that are delinquent (late).

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?
Investors seem to favor lower Debt to Income Ratios, as well as higher Cash 
Flows and higher ProsperScores. This would suggest that investors that
participate on Prosper are usually selective. Also, there doesn't seem to be any
significant association between incomes being verifiable and ability to pay a 
loan off or not.

### What was the strongest relationship you found?
At the moment, the strongest relationship I've found is between the number of 
investors and a borrower's ProsperScore. I'm sure I will find stronger 
relationships moving forward.

# Multivariate Plots Section

I've looked at a variety of factors at this point. It's time to combine and 
compare them. Let's see how investors stack up against Prosper in their analysis
of loans. I'll look at median numbers of investors, ProsperScores, and whether
loans were Completed or Defaulted:

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots}
# Investors vs Prosper Score vs Loan Status
filtered_data <- subset(loan_data, 
                        LoanStatus == "Completed" | LoanStatus == "Defaulted")

ggplot(aes(x = ProsperScore, y = Investors), data = filtered_data) +
  geom_point(aes(color = LoanStatus), stat = 'summary', fun.y = median) +
  ylab("Median Investors") +
  geom_smooth(aes(color = LoanStatus), stat = 'summary', fun.y = median, 
              size=0.5) +
  scale_x_continuous(breaks = seq(0, 10, 1), limits = c(0, 10))
```

As shown before, the number of investors drastically increases as ProsperScore
increases. But what is worrisome is that at ProsperScores between 5.5 and 8.5,
signifnicantly more investors funded loans that ended up Defaulted rather than
Completed. The risks associated with loans with scores of 5-6 are reasonable,
but why did so many investors fund failed loans at scores 7-8.5? Was Prosper 
inaccurate in their risk scores, or were investors wrong in their analysis of 
the loan, or both? It's difficult to tell as of now. Let's keep digging in other
directions.

Next let's see how accurate investors were in their analysis of debt to income
ratios, and of how their loans would turn out.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Debt to Income ratios vs number of investors lending vs Loan Status
filtered_data <- subset(loan_data, 
                        LoanStatus != "Current" & LoanStatus != "Cancelled" & 
                          LoanStatus != "Chargedoff")

ggplot(aes(x = DebtToIncomeRatio, y = Investors, color = LoanStatus), 
       data = filtered_data) +
  geom_point(alpha = 0.5, size = 1.0, position = "jitter") +
  scale_x_continuous(breaks = seq(0, 1.0, 0.05), limits = c(0, 1.0)) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
```

The vast majority of loans are completed, fortunately. Once Debt to Income Ratio
hits 0.55 and above, we see a growing proportion of Defaulted loans to 
Completed loans, and we also see the number of investors decreasing, with a
couple outliers in the 0.9 to 1.0 range. (Very nice lenders they must be, or 
maybe they charged huge interest). So for the most part, investors are making
the right calls when it comes to Debt to Income Ratios.

I would hope that investors are not discriminating based on income ranges. What
should matter is net cash flow and debt, not income. Let's see whether investors
are discriminating or not based on income:

```{r message=FALSE, warning=FALSE}
# Income vs Cash Flow vs Investors
# first need to bucket cash flow so I can categorize it
loan_data$CashFlow.bucket <- cut(loan_data$CashFlow, 
                                 c(-1200, 800, 2800, 4800, 6800, 10000, 
                                   2000000))

ggplot(aes(x = IncomeRange, y = Investors), data = loan_data) +
  geom_boxplot() +
  facet_wrap(~CashFlow.bucket) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  coord_cartesian(ylim = c(0, 500))
```

The pattern I am seeing here is that for low to medium Cash Flows, all the 
way from -1200 to 4800, investor numbers are clearly higher for higher incomes. 
But once we hit the high to elite cash flow levels, 4800+ to 2M, investor 
numbers don't show a clear preference for higher incomes. So it seems that some 
income discrimination exists, but only for low to medium Cash Flows, not for 
high to elite Cash Flows.

Let's next confirm that my earlier assumption is correct: that debt and cash
flow together are good indicators of being able to pay off a loan.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Debt to Income Ratio vs Cash Flow vs Loan Status
filtered_data <- filtered_data <- subset(loan_data, 
                        LoanStatus == "Completed" | LoanStatus == "Defaulted")

ggplot(aes(x = DebtToIncomeRatio, y = CashFlow, color = LoanStatus), 
       data = filtered_data) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  scale_x_continuous(breaks = seq(0, 5, 0.25), limits = c(0, 5)) +
  scale_y_continuous(breaks = seq(0, 10000, 500), limits = c(0, 10000))
```

Once again, let's not look at total amounts of Completed and Defaulted loans,
because there are far more Completed than Defaulted overall. Let's look at 
proportions. Higher CashFlows and lower DebtToIncome ratios are clearly 
associated with much higher proportions of Completed loans, while low CashFlows
and high DebtToIncome ratios are associated with much more closer proportions of
Completed vs Defaulted loans. So I was correct in this assumption.

Lastly, I want to take a look at the terms of the loans itself. Do factors such
as interest rate (called LenderYield here) or length of loan (called Term here)
predict the difficulty of the borrower paying off the loan?

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Converting Term to an ordered factor instead of continuous 
loan_data$Term <- factor(loan_data$Term, levels=c(12,36,60), ordered=TRUE)

# LenderYield (interest rate) vs Term (loan length) vs Loan Status
filtered_data <- subset(loan_data, 
                        LoanStatus != "Current" & LoanStatus != "Cancelled" & 
                          LoanStatus != "Chargedoff")

ggplot(aes(x = Term, y = LenderYield), data = filtered_data) +
  geom_jitter(alpha = 0.1) +
  facet_wrap(~LoanStatus)
```

So right away we can see that there is only a small number of available loan
Terms. There is one noticable pattern for loan Term: for Defaulted loans, a 3
year loan was the most common. As for LenderYield, there's no discernible 
pattern. It looks like for every Loan Status, whether Complete, Defaulted, or 
Delinquent, there is a huge variety of interest rates on these loans.

Now that I have explored quite a few variables, I want to create a model and try
predicting a potential Loan Status for it. At this moment I do not know how to
build a model of a categorical variable, so I will filter the data and set a
Defaulted Loan equal to 100 and a Completed Loan equal to 0.

```{r message=FALSE, warning=FALSE}
# filter data first
filtered_data <- filtered_data <- subset(loan_data, 
                        LoanStatus == "Completed" | LoanStatus == "Defaulted")

# change Completed to 0 and Defaulted to 1
filtered_data$ModelData[filtered_data$LoanStatus == "Completed"] <- 0
filtered_data$ModelData[filtered_data$LoanStatus == "Defaulted"] <- 100

m1 <- lm(ModelData ~ I(CashFlow/1000), data = filtered_data)
m2 <- update(m1, ~ . + DebtToIncomeRatio)
m3 <- update(m2, ~ . + ProsperScore)
mtable(m1, m2, m3)
```

With CashFlow, DebtToIncomeRatio, and ProsperScore all at zero, the LoanStatus 
can be expected to be at 13.011. (In this crude model, a 0 represents a 
Completed loan and a 100 represents a Defaulted loan.) For $1000 increase in 
CashFlow, we can expect the LoanStatus to drop by -0.274. For every unit
increase in DebtToIncomeRatio on top of that, we can expect at 2.052 unit 
increase in LoanStatus, inching closer to a Default. And if we add a unit
increase of ProsperScore, we see a -0.747 drop back down towards a Completed
loan.

Now let's try to predict a LoanStatus with some artificial data.

```{r message=FALSE, warning=FALSE}
# First, let's create a simple dataframe with one row of artificial data.
myLoan = data.frame(CashFlow = 2000, DebtToIncomeRatio = 0.5, ProsperScore = 5)
# Then, let's invoke the predict() function and pass in our most updated model 
# object, m3, our new data frame, and our confidence interval, 0.95.
modelEstimate = predict(m3, newdata = myLoan, interval = "prediction", 
                        level = 0.95)
modelEstimate
```

With a Cashflow of 2000, a DebtToIncomeRatio of 0.5, and a ProsperScore of 5,
our model predicts a LoanStatus 0f 7.14, which is pretty close to a Completed
Loan value. The lower and upper bounds have quite a big range, though. This 
model is obviously very crude.


# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?
Clearly, high ProsperScores given by Prosper inspire more investors to lend.
Even for loans that ended up defaulting, a relatively high median number of
investors were quick to lend when ProsperScores were around 8. On the other 
hand, investors made good decisions when it came to Debt to Income Ratios. They
were much more conservative when lending here and it paid off; a higher
proportion of loans defaulted with higher Debt to Income Ratios, while a lower
number of investors jumped on board. Income discrimination also seems to be 
present, but not ubiquitously. Only when a borrower has a low to medium cash
flow, more investors flock to higher income borrowers. But when a borrower has
a high to elite cash flow, investors don't necessarily prefer any income. Cash
Flow and Debt to Income Ratios are also strongly associated with ability to pay
off loans, as expected. The lower the Cash Flow and higher the Debt to Income
Ratio, the higher the proportion of defaulted loans.

### Were there any interesting or surprising interactions between features?
The two biggest surprises to me were: 1) that loan length (Term) and interest
rate (LenderYield) had no significant correlation or effect on the borrower's
ability to pay the loan off, and 2) that the verifiability/authenticity of a
borrower's listed income had no noticeable effect on the borrower's ability to
pay the loan off (this was in the bivariate section, but surprising 
nonetheless).

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.
I did try to create a linear model. It was very crude; the details are above. 
The strength is that I included multiple variables I found significant, but the
limitation is that I forced a linear model on a categorical variable by slightly
transforming the LoanStatus variable.
------


# Final Plots and Summary

### Plot One
```{r message=FALSE, warning=FALSE, echo=FALSE, Plot_One}
# Loan statuses vs Prosper ratings
filtered_data <- subset(loan_data, ProsperRating..Alpha. != "" & 
                          LoanStatus != "Current")

ggplot(aes(x = LoanStatus), data = filtered_data) +
  geom_histogram(aes(fill = filtered_data$ProsperRating..Alpha.), 
                 position = "dodge", stat = "count") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1),
        plot.title = element_text(hjust = 0.5)) +
  ggtitle("Loan Statuses vs Prosper Ratings") +
  xlab("Loan Status") +
  ylab("Count") +
  guides(fill = guide_legend(title = "Prosper Rating (AA - HR)"))
```
```{r message=FALSE, warning=FALSE, echo=FALSE, Plot_OneB}
# Loan statuses vs Prosper ratings
defaulted = subset(loan_data, 
                   LoanStatus == 'Defaulted' & ProsperRating..Alpha. != '')

ggplot(aes(x = LoanStatus), data = defaulted) +
  geom_bar(aes(fill = defaulted$ProsperRating..Alpha.), 
                 position = "dodge", stat = "count") +
  theme(plot.title = element_text(hjust = 0.5)) +
  ggtitle("Defaulted Loans vs Prosper Ratings") +
  xlab("Loan Status") +
  ylab("Count") +
  guides(fill = guide_legend(title = "Prosper Rating (AA - HR)"))
```

### Description One
I did two plots here in place of one, because to tell the full story I needed a
second plot to zoom into Defaulted Loans. For loans that were completely paid in
full, Prosper was not very accurate with their scores. For completed loans, 
Prosper had given a similar amount of good ratings (AA, A, B), as bad ratings 
(C, D, E). For defaulted loans, Prosper was fortunately more accurate in their 
pre-ratings. The highest rating, AA, was given the least. Bad ratings (C, D, E) 
were given in much higher proportions than the best 3 ratings.

So it seems Prosper takes the conservative route and hands out bad ratings at
the slighest notion of a potential defaulted loan in the future.

### Plot Two
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}
# Debt to Income Ratio vs Cash Flow vs Loan Status
filtered_data <- filtered_data <- subset(loan_data, 
                        LoanStatus == "Completed" | LoanStatus == "Defaulted")

ggplot(aes(x = DebtToIncomeRatio, y = CashFlow, color = LoanStatus), 
       data = filtered_data) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1),
        plot.title = element_text(hjust = 0.5)) +
  scale_x_continuous(breaks = seq(0, 5, 0.25), limits = c(0, 5)) +
  scale_y_continuous(breaks = seq(0, 10000, 500), limits = c(0, 10000)) +
  ggtitle("Debt to Income Ratio vs Cash Flow vs Loan Status") +
  xlab("Debt to Income Ratio (0-5)") +
  ylab("Cash Flow ($ USD)") +
  guides(fill = guide_legend(title = "Loan Status"))
```

### Description Two
Let's not look at total amounts of Completed and Defaulted loans, because there 
are far more Completed than Defaulted overall. Let's look at proportions. Higher
CashFlows and lower DebtToIncome ratios are clearly associated with much higher
proportions of Completed loans, while low CashFlows and high DebtToIncome ratios
are associated with much more closer proportions of Completed vs Defaulted 
loans. Both of these trends can be used by investors and Prosper to analyze
borrowers' characteristics before loans are given.

### Plot Three
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Three}
# Income vs Cash Flow vs Investors
# first need to bucket cash flow so I can categorize it
loan_data$CashFlow.bucket <- cut(loan_data$CashFlow, 
                                 c(-1200, 800, 2800, 4800, 6800, 10000, 
                                   2000000))
levels(loan_data$CashFlow.bucket) <- c("-$1200 to $800", "$800 to $2800",
                                       "$2800 to $4800", "$4800 to $6800",
                                       "$6800 to $10000", "$10000 to $2000000")

ggplot(aes(x = IncomeRange, y = Investors), data = loan_data) +
  geom_boxplot() +
  facet_wrap(~CashFlow.bucket) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1),
        plot.title = element_text(hjust = 0.5)) +
  coord_cartesian(ylim = c(0, 500)) +
  ggtitle("Income vs Investors by Cash Flow Range") +
  xlab("Income Range ($ USD)") +
  ylab("Number of Investors")
```

### Description Three
The pattern I am seeing here is that for low to medium Cash Flows, all the 
way from -1200 to 4800, investor numbers are clearly higher for higher incomes. 
But once we hit the high to elite cash flow levels, 4800+ to 2M, investor 
numbers don't show a clear preference for higher incomes. So it seems that some 
income discrimination exists, but only for low to medium Cash Flows, not for 
high to elite Cash Flows.
------


# Reflection
Getting the chance to explore this dataset gave me very good practice with the
concepts I've learned in Exploratory Data Analysis. I encountered a few
struggles, including: overplotting, missing data ("N/A"), incorrect data formats
(dates), unnecessary variables & factors, and most of all, trying to build a
model of a categorical variable. But for the most part, I think my analysis went
well. I was able to explore quite a few variables and see their relationships
with one another. I was able to use a variety of plot types in meaningful ways.
A couple things that surprised me: that loan length and interest rate had no 
significant correlation or effect on the borrower's ability to pay the loan off,
and that the verifiability/authenticity of a borrower's listed income had no 
noticeable effect on the borrower's ability to pay the loan off. As for future
work, I really need to learn how to build an accurate prediction model for Loan 
Status, instead of changing it to a continuous variable to force a linear model 
to work.